---
title: "Visualizations of the multilayer dynamic networks"
date: "2026-08-20"
output:
  html_document:
    theme: default
    highlight: kate
    toc: true
    toc_float: true
    number_sections: false
    smooth_scroll: true
---

<style>
/* ancho del TOC */
#TOC {
  width: 200px;
}

/* elimina separación extra */
.toc-content {
  padding-left: 0px !important;
}

/* contenedor principal */
.main-container {
  max-width: 1200px;
}
</style>

# Data preparation

```{r message=FALSE, warning=FALSE}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                           0.1 Library loading                          ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
library(igraph)
library(ggraph)
library(tidygraph)
library(dplyr)
library(scales)
library(network)
library(networkDynamic)
library(ndtv)
```


```{r}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                            0.2 Data loading                            ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
load("icews_tensor_proyecto.RData")

n_tiempos <- dim(tensor_array)[4]
n_capas <- dim(tensor_array)[3]
n_paises <- dim(tensor_array)[1]

nombres_capas <- c("Material-", "Material+", "Verbal-", "Verbal+")
nombres_paises <- dimnames(tensor_array)$Source
```

```{r}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                 1. Select the 25 most central countries                ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

# Array to store centrality measures
page_cent <- array(NA,
                  dim = c(n_paises, n_capas, n_tiempos),
                  dimnames = list(nombres_paises, paste0("Layer_",1:n_capas), 
                                  paste0("Time_",1:n_tiempos)))

# Calculate Page Rank centrality for every graph
for(m in 1:n_capas){
  for(t in 1:n_tiempos){
    A <- tensor_array[,,m,t]
    g <- graph_from_adjacency_matrix(A, mode = "directed", weighted = TRUE, diag = FALSE)
    page_cent[,m,t] <- page_rank(g, directed = TRUE, weights = E(g)$weight)$vector
  }
}

# Average centrality measures across time
cent_mean <- apply(page_cent, c(1,2), mean, na.rm = TRUE)

# Identify the top 25 most central countries
top25_layers <- vector("list", n_capas)
for(m in 1:n_capas){
  ord <- order(cent_mean[,m], decreasing = TRUE)
  top25_layers[[m]] <- data.frame(Country = rownames(cent_mean)[ord[1:25]],
                                  MeanEigenCentrality = cent_mean[ord[1:25], m])
}
top25_names <- lapply(top25_layers, \(x) x$Country)

# Countries that are among the top 25 in all layers 
Reduce(intersect, top25_names) 
# There are 19 countries

# - - - - - - - - - - - - - - - - Gloabl top 25 - - - - - - - - - - - - - - - - 

# Compute the average eigenvector centrality across all time points
overall_mean <- rowMeans(cent_mean)

# Rank countries in decreasing order according to their average centrality
ord <- order(overall_mean, decreasing = TRUE)

# Select the 25 countries with the highest average eigenvector centrality
top25_global <- data.frame(Country = names(overall_mean)[ord[1:25]],
                           MeanEigenCentrality = overall_mean[ord[1:25]])
# The top 25 include the 19 countries identified previously
# plus 6 additional countries

# Store the names of the final set of selected countries
paises_final <- top25_global$Country

# Obtain the indices of the selected countries in the original country list
indices <- match(paises_final, nombres_paises)
```

# Visualization functions

The following functions are used to avoid repeating the same graph construction, formatting, and dynamic-network code for each layer.


```{r}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
#                          2.1 Static network function                     
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

create_static_network <- function(tensor_array,
                                  indices,
                                  layer,
                                  time,
                                  label_map = c()) {
  
  # Extract the adjacency matrix for the selected layer and time
  A <- tensor_array[indices, indices, layer, time]
  
  # Construct a directed weighted graph from the adjacency matrix
  g <- graph_from_adjacency_matrix(A, mode = "directed", weighted = TRUE, diag = FALSE)
  
  # Compute the degree of each node, considering both incoming and outgoing edges
  V(g)$degree <- igraph::degree(g, mode = "all")
  
  # Compute the node coordinates using the Nicely layout algorithm
  set.seed(9)
  layout <- layout_nicely(g)
  
  # Scale node size
  V(g)$size <- rescale(V(g)$degree, to = c(5, 15))
  
  # Scale edge width
  E(g)$width <- rescale(E(g)$weight, to = c(2, 6))
  
  # Define labels only for selected countries
  V(g)$label <- ifelse(V(g)$name %in% names(label_map), unname(label_map[V(g)$name]),
                       NA_character_)
  
  return(
    list(graph = g, layout = layout)
  )
}

# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
#                   2.2 Static network plotting function               
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

plot_static_network <- function(network_object,
                                title = NULL,
                                layout = NULL) {
  
  g <- network_object$graph
  if (is.null(layout)) {
    layout <- network_object$layout
  }
  
  plot(g,
       layout = layout,
    
      # Node settings
      vertex.color = "dodgerblue2",
      vertex.frame.color = "white",
      vertex.size = V(g)$size,
    
      # Label settings
      vertex.label = V(g)$label,
      vertex.label.color = "black",
      vertex.label.cex = 1,
    
      # Edge settings
      edge.width = E(g)$width,
      edge.color = adjustcolor("grey60", alpha.f = 0.5),
      edge.arrow.size = 0.4,
    
      # Do not display a legend
      rescale = TRUE
      )
}

# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
#                         2.3 Static comparison function                    
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

plot_static_comparison <- function(tensor_array,
                                   indices,
                                   layer,
                                   layer_name,
                                   times = c(1, 132),
                                   label_maps,
                                   layout_reference_time = times[1]) {
  
  # Create the networks for the selected time points
  networks <- lapply(
    times,
    function(time) {
      create_static_network(tensor_array = tensor_array,
                            indices = indices,
                            layer = layer,
                            time = time,
                            label_map = label_maps[[as.character(time)]])
    }
  )
  
  # Identify the network corresponding to the reference time
  reference_index <- match(layout_reference_time, times)
  
  # Use the layout from the reference time for all time points
  common_layout <- networks[[reference_index]]$layout
  
  # Set the graphical parameters
  par(mfrow = c(1, length(times)), mar = c(0, 0, 2, 0) )
  
  # Plot each network
  for (i in seq_along(networks)) {
    plot_static_network(network_object = networks[[i]],
                        layout = common_layout,
                        title = paste0(layer_name, " network: t = ", times[i])
                        )
  }
}
```



```{r}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
#                            2.4 Dynamic network function                    
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

create_dynamic_network <- function(tensor_array,
                                   indices,
                                   layer,
                                   times,
                                   country_names) {
  
  # - - - - - - - - - - - - Compute global edge weights - - - - - - - - - - - - 
  
  # Extract all edge weights across the selected time points
  pesos_todos <- unlist(
    lapply(times,
           function(t) {
             A <- tensor_array[indices, indices, layer, t]
             A[A != 0 & is.finite(A)]
             }
           ) )
  
  # Remove any remaining non-finite values
  pesos_todos <- pesos_todos[is.finite(pesos_todos)]
  
  # Compute the global minimum and maximum edge weights
  min_peso <- min(pesos_todos)
  max_peso <- max(pesos_todos)
  
  
  # - - - - - - - - - - - - - Construct the edge spells - - - - - - - - - - - - 
  
  edge_spells <- lapply(times,
                        function(t) {
      # Extract the adjacency matrix at time t
      A <- tensor_array[indices, indices, layer, t]
      
      # Identify the existing edges in the adjacency matrix
      ind <- which(A != 0 & is.finite(A), arr.ind = TRUE)
      
      # Extract edge information
      data.frame(onset = t,
                 terminus = t + 1,
                 tail = ind[, 1],
                 head = ind[, 2],
                 weight = A[ind],
                 edge.lwd = if (max_peso == min_peso) {
                   rep(5, length(ind))} else {
                     1 + 9 * ((A[ind] - min_peso) / (max_peso - min_peso) )
                     }
                 )
      }
  )
  
  # Remove empty time points
  edge_spells <- Filter(Negate(is.null), edge_spells)
  
  # Combine all edge spells into a single data frame
  edge_spells <- do.call(rbind, edge_spells)
  
  # - - - - - - - - - - - - - Create the dynamic network - - - - - - - - - - - -
  
  samp.dyn <- networkDynamic(edge.spells = edge_spells,
                             create.TEAs = TRUE,
                             edge.TEA.names = c("weight", "edge.lwd"),
                             directed = TRUE)
  
  # Assign country names to the vertices
  network.vertex.names(samp.dyn) <- country_names
  
  return(samp.dyn)
}

# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                 2.5 Dynamic network rendering function            
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

render_dynamic_network <- function(samp.dyn,
                                   layer_name,
                                   label_map = c(),
                                   label_cex = 0.8) {
  
  # Retrieve the vertex names
  nombres <- network.vertex.names(samp.dyn)
  
  # Display labels only for selected countries
  labels_seleccionados <- ifelse(nombres %in% names(label_map),
                                 unname(label_map[nombres]), "")
  
  # Set the seed to ensure reproducibility of the network animation
  set.seed(123)
  
  # Render the dynamic network as an interactive HTML widget
  render.d3movie(samp.dyn,
                 plot.par = list(displaylabels = TRUE,
                                 label = labels_seleccionados,
                                 label.cex = label_cex,
                                 main = paste0(layer_name, " network (2004-2005)"),
                                 vertex.border = "white",
                                 edge.col = "grey60",
                                 edge.lwd = "edge.lwd"),
                 output.mode = "htmlWidget")
}
```

# Visualization settings

```{r}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
#                         3.1 Visualization settings                    
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

# Define the time points used in the dynamic networks
tiempos <- 1:24

# Extract the names of the selected countries
nombres_paises <- dimnames(tensor_array)[[1]][indices]

# Define the layers
network_layers <- data.frame(layer = 1:4,
                             name = nombres_capas)

# Define labels for each static visualization
static_labels <- list(
  
  "Material-" = list(`1` = c("United States" = "USA",
                             "Iraq" = "IRQ",
                             "Israel" = "ISR"),
                     `132` = c("United States" = "USA",
                               "Iraq" = "IRQ",
                               "Syria" = "SYR")
                     ),
  
  "Material+" = list(`1` = c("United States" = "USA",
                             "Iraq" = "IRQ"),
                     `132` = c("United States" = "USA",
                               "Iraq" = "IRQ")
                     ),
  
  "Verbal-" = list(`1` = c("United States" = "USA",
                           "Iraq" = "IRQ",
                           "Israel" = "ISR"),
                   `132` = c("United States" = "USA",
                             "Iraq" = "IRQ",
                             "Russian Federation" = "RUS")
                   ),
  
  "Verbal+" = list(`1` = c("United States" = "USA",
                           "United Kingdom" = "UK"),
                   `132` = c("United States" = "USA",
                             "United Kingdom" = "UK")
  )
)

# Define labels for the dynamic networks
dynamic_labels <- list(
  "Material-" = c("United States" = "USA", "Iraq" = "IRQ"),
  
  "Material+" = c("United States" = "USA", "Iraq" = "IRQ"),
  
  "Verbal-" = c("United States" = "USA", "Iraq" = "IRQ"),
  
  "Verbal+" = c("United States" = "USA", "United Kingdom" = "UK")
)
```

# Visualizations

To identify the most relevant actors in the network, the three countries with the highest degree centrality are highlighted. In all static network visualizations, node size is proportional to degree centrality.

## Material- type network

```{r fig.height=6, fig.width=12, message=FALSE, warning=FALSE, out.width="100%"}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                         Material- static network                    ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

plot_static_comparison(tensor_array = tensor_array,
                       indices = indices,
                       layer = 1,
                       layer_name = "Material-",
                       times = c(1, 132),
                       label_maps = static_labels[["Material-"]],
                       layout_reference_time = 1)
```


```{r message=FALSE, warning=FALSE, out.width="100%"}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                        Material- dynamic network                    ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

dynamic_material_minus <- create_dynamic_network(tensor_array = tensor_array,
                                                 indices = indices,
                                                 layer = 1,
                                                 times = tiempos,
                                                 country_names = nombres_paises)

render_dynamic_network(samp.dyn = dynamic_material_minus,
                       layer_name = "Material-",
                       label_map = dynamic_labels[["Material-"]],
                       label_cex = 0.8)
```

## Material+ type network


```{r fig.height=6, fig.width=12, message=FALSE, warning=FALSE, out.width="100%"}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                         Material+ static network                    ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

plot_static_comparison(tensor_array = tensor_array,
                       indices = indices,
                       layer = 2,
                       layer_name = "Material+",
                       times = c(1, 132),
                       label_maps = static_labels[["Material+"]],
                       layout_reference_time = 1)
```


```{r message=FALSE, warning=FALSE, out.width="100%"}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                        Material+ dynamic network                    ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

dynamic_material_plus <- create_dynamic_network(tensor_array = tensor_array,
                                                indices = indices,
                                                layer = 2,
                                                times = tiempos,
                                                country_names = nombres_paises)

render_dynamic_network(samp.dyn = dynamic_material_plus,
                       layer_name = "Material+",
                       label_map = dynamic_labels[["Material+"]],
                       label_cex = 0.8)
```

## Verbal- type network

```{r fig.height=6, fig.width=12, message=FALSE, warning=FALSE, out.width="100%"}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                          Verbal- static network                     ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

plot_static_comparison(tensor_array = tensor_array,
                       indices = indices,
                       layer = 3,
                       layer_name = "Verbal-",
                       times = c(1, 132),
                       label_maps = static_labels[["Verbal-"]],
                       layout_reference_time = 132)
```

### Dynamic network

```{r message=FALSE, warning=FALSE, out.width="100%"}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                         Verbal- dynamic network                     ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

dynamic_verbal_minus <- create_dynamic_network(tensor_array = tensor_array,
                                               indices = indices,
                                               layer = 3,
                                               times = tiempos,
                                               country_names = nombres_paises)

render_dynamic_network(samp.dyn = dynamic_verbal_minus,
                       layer_name = "Verbal-",
                       label_map = dynamic_labels[["Verbal-"]],
                       label_cex = 0.8)
```

## Verbal+ type network

```{r fig.height=6, fig.width=12, message=FALSE, warning=FALSE, out.width="100%"}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                          Verbal+ static network                     ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

plot_static_comparison(tensor_array = tensor_array,
                       indices = indices,
                       layer = 4,
                       layer_name = "Verbal+",
                       times = c(1, 132),
                       label_maps = static_labels[["Verbal+"]],
                       layout_reference_time = 132)
```

```{r message=FALSE, warning=FALSE, out.width="100%"}
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
####                         Verbal+ dynamic network                     ####
# ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

dynamic_verbal_plus <- create_dynamic_network(tensor_array = tensor_array,
                                              indices = indices,
                                              layer = 4,
                                              times = tiempos,
                                              country_names = nombres_paises)

render_dynamic_network(samp.dyn = dynamic_verbal_plus,
                       layer_name = "Verbal+",
                       label_map = dynamic_labels[["Verbal+"]],
                       label_cex = 0.8)
```